---
############################################################
# Lead exposure and school readiness
#
# R script: spline and figure.Rmd
#
# Purpose:
# Evaluate non-linearity using natural cubic splines. Prepare figure 2
#
# Input:
# analysis_dataset.dta
#
# Output:
# Spline model and figures
#
# Software:
# R 4.3.3
############################################################
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r}
##Load libraries
library(tidyverse)
library(lme4)
library(splines)
library(ggplot2)
library(dplyr)
library(readr)
library(haven)
library(splines)
library(ggeffects)
library(sjPlot)
library(scales) 
```


```{r}
##Set directory
setwd("C:/Users/AditiRoy/OneDrive - Centre For Chronic Disease Control/ECD_Urban Pollution/pb_ECD_paper")

```

## Read data
```{r}
data1 <- read_dta("analysis_dataset.dta")
```

##Convert factors
```{r}
data1$record_id <- as.factor(data1$record_id)
data1$m_edu_recode3 <- as.factor(data1$m_edu_recode3)

data1 <- data1 %>%  
  mutate(
    log2bpb2_c = scale(log2bpb2, center = TRUE, scale = FALSE),
    c_age_c    = scale(c_age, center = TRUE, scale = FALSE),
    crowding_c = scale(crowding, center = TRUE, scale = FALSE)
  )
```

  

```{r}
#reshape the data 
library(tidyr)

data1 <- data1 %>%
  pivot_longer(
    cols = c(
      z_theta_total2m_t1, z_theta_total2m_t2,
      z_score_mrant_t1, z_score_mrant_t2
    ),
    names_to = c(".value", "time"),
    names_pattern = "(.*?)([12])$"
  )

data1$time <- as.numeric(data1$time)

```


##Fit Linear mixed model (Pb and IRT-derived IDEA_z scores)
```{r}

model_linear <- lmer(
  z_theta_total2m_t ~ 
    log2bpb2 +
    time +
    c_age +
    child_hb +
    crowding +
    factor(m_edu_recode3) +
    (1 | record_id),
  data = data1,
  REML = TRUE
)
summary(model_linear)
est <- fixef(model_linear)["log2bpb2"]
se  <- summary(model_linear)$coefficients["log2bpb2", "Std. Error"]
ci  <- confint(model_linear, parm = "log2bpb2", method="Wald")

cat("Beta:", round(est,3),
    "SE:", round(se,3),
    "95% CI:", round(ci[1],3), "to", round(ci[2],3))

```

## Fit Spline model (Pb and IRT-derived IDELA-z-scores)

```{r}
data1$log2bpb2_c <- scale(data1$log2bpb2, center = TRUE, scale = FALSE)
model_spline <- lmer(
  z_theta_total2m_t ~ 
    ns(log2bpb2_c, df = 3) +
    time +
    c_age +
    child_hb +
    crowding +
    m_edu_recode3 +
    (1 | record_id),
  data = data1,
  REML = FALSE,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 200000)
)
)

summary(model_spline)
# Find the mean that was used to center the variable
lead_mean <- attr(data1$log2bpb2_c, "scaled:center")

```

##Compare the generalized linear model with spline model
```{r}
anova(model_linear, model_spline)
```


##Testing interactions- Pb and child_hb on IRT-derived IDELA-z scores
```{r}
data1$child_hb_c <- scale(data1$child_hb, center = TRUE, scale = FALSE)

model_linear_int <- lmer(
  z_theta_total2m_t ~ 
    log2bpb2 * child_hb_c +
    time +
    c_age +
    crowding +
    factor(m_edu_recode3) +
    (1 | record_id),
  data = data1,
  REML = FALSE
)
summary(model_linear_int)
```

```{r}
hb_mean <- mean(data1$child_hb, na.rm = TRUE)
hb_sd   <- sd(data1$child_hb, na.rm = TRUE)

hb_levels <- c(hb_mean - hb_sd, hb_mean, hb_mean + hb_sd)

# Generate predictions
pred <- ggpredict(
  model_linear_int,
  terms = c(
    "log2bpb2 [all]",
    "child_hb_c [-1,0,1]"
  )
)
```

## Create interaction plot (supplementary information)
```{r}
ggplot(
  pred,
  aes(
    x = x,
    y = predicted,
    linetype = group
  )
) +
 # 95% confidence intervals
  geom_ribbon(
    aes(
      ymin = conf.low,
      ymax = conf.high,
      group = group
    ),
    fill = "grey70",
    alpha = 0.25,
    colour = NA
  ) +

  # Prediction lines
  geom_line(
    linewidth = 1,
    colour = "black"
  ) +

  # Hb labels
  scale_linetype_manual(
    values = c("longdash", "solid", "dotted"),
    labels = c(
      "Low Hb (10.0 g/dL)",
      "Mean Hb (11.3 g/dL)",
      "High Hb (12.6 g/dL)"
    )
  ) +
  guides(
  linetype = guide_legend(
    nrow = 2,
    byrow = TRUE
  )
) +

  # Convert log2 BPb back to actual BPb values
  scale_x_continuous(
    breaks = log2(c(4.5, 11.1, 20, 40)),
    labels = c("4.5", "11.1", "20", "40")
  ) +

  labs(
  x = "Blood lead level (μg/dL)",
  y = "Predicted IDELA total z-score",
  linetype = "Hemoglobin level"
  ) +

  annotate(
    "text",
    x = log2(20),
    y = 0.05,
    label = "P-interaction = 0.05",
    size = 4
  ) +

  theme_classic(base_size = 12) +
  theme(
  axis.text.y = element_text(size = 7)
) +

  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )

# Save figure
ggsave(
  "IDELA_Hb_interaction_curve.tiff",
  width = 7,
  height = 7,
  dpi = 600,
  units = "in",
  compression = "lzw"
)
```


##Figure 2A: Forest plot of the estimated difference in IDELA z-score

```{r}
# # comparing the 90th vs 10th percentile of blood lead level,
# stratified by child hemoglobin. Estimates were obtained from
# the primary mixed-effects model fitted in STATA.

contrast_df <- data.frame(
  Hb_level = c(
    "Low Hb (10.0)",
    "Mean Hb (11.3)",
    "High Hb (12.6)"
  ),
  estimate = c(-0.27, -0.10, 0.06),
  lower = c(-0.49, -0.27, -0.19),
  upper = c(-0.04, 0.07, 0.31)
)

contrast_df$Hb_level <- factor(
  contrast_df$Hb_level,
  levels = c(
    "Low Hb (10.0)",
    "Mean Hb (11.3)",
    "High Hb (12.6)"
  )
)

ggplot(contrast_df, aes(x = Hb_level, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.15
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  theme_classic() +
  labs(
    x = "Child hemoglobin level (g/dL)",
    y = "Difference in IDELA z-score\n(90th vs 10th percentile BPb)"
  ) +
  annotate(
    "text",
    x = 2,
    y = 0.42,
    label = "P-interaction = 0.05",
    size = 4
  )
#Save figure
ggsave("Figure2_IDELA_Hb_interaction.tiff",
       width = 6,
       height = 5,
       units = "in",
       dpi = 600,
       compression = "lzw")

ggsave("Figure2_IDELA_Hb_interaction.pdf",
       width = 6,
       height = 5,
       units = "in")

```

##Fit linear model with WM
```{r}
model_WM_lin1 <- lmer(
 z_score_mrant_t ~ 
    log2bpb2 +
    time +
    c_age +
    child_hb +
    crowding +
    factor(m_edu_recode3) +
    (1 | record_id),
  data = data1,
  REML = FALSE
)
summary(model_WM_lin1)
est <- fixef(model_WM_lin1)["log2bpb2"]
se  <- summary(model_WM_lin1)$coefficients["log2bpb2", "Std. Error"]
ci  <- confint(model_WM_lin1, parm = "log2bpb2", method="Wald")

cat("Beta:", round(est,3),
    "SE:", round(se,3),
    "95% CI:", round(ci[1],3), "to", round(ci[2],3))

```

##Spline with WM
```{r}
model_spl_mrantscore <- lmer(
  z_score_mrant_t ~ 
    ns(log2bpb2_c, df = 3) +
    time +
    c_age +
    child_hb +
    crowding +
    m_edu_recode3 +
    (1 | record_id),
  data = data1,
  REML = FALSE,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 200000)
)
)

summary(model_spl_mrantscore)
```

##Compare the generalized linear model with spline model (WM)
```{r}
anova(model_WM_lin1, model_spl_mrantscore)

```


##Figure 2B: Forest plot of the estimated difference in Working Memory score
```{r}
# Working memory contrasts
contrast_df_wm <- data.frame(
  Hb_level = c(
    "Low Hb (10.0)",
    "Mean Hb (11.3)",
    "High Hb (12.6)"
  ),
  estimate = c(
    -0.24394906,
    -0.10740147,
     0.02914611
  ),
  lower.CL = c(
    -0.4441881,
    -0.2742444,
    -0.1875065
  ),
  upper.CL = c(
    -0.04370998,
     0.05944149,
     0.24579877
  )
)

contrast_df_wm$Hb_level <- factor(
  contrast_df_wm$Hb_level,
  levels = c(
    "Low Hb (10.0)",
    "Mean Hb (11.3)",
    "High Hb (12.6)"
  )
)

# Plot
p <- ggplot(
  contrast_df_wm,
  aes(x = Hb_level, y = estimate)
) +

  geom_point(
    size = 3,
    color = "black"
  ) +

  geom_errorbar(
    aes(
      ymin = lower.CL,
      ymax = upper.CL
    ),
    width = 0.15,
    color = "black"
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +

  coord_cartesian(
    ylim = c(-0.5, 0.5)
  ) +

  scale_y_continuous(
    breaks = seq(-0.5, 0.5, by = 0.1)
  ) +

  theme_classic(base_size = 12) +

  labs(
    x = "Child hemoglobin level (g/dL)",
    y = "Difference in working memory z-score\n(40 vs 4.5 μg/dL BPb)"
  ) +

  annotate(
    "text",
    x = 2,
    y = 0.42,
    label = "P-interaction = 0.03",
    size = 4
  )

# Display plot
p

# Save high-resolution TIFF
ggsave(
  filename = "BPb_Hb_interaction_WM.tiff",
  plot = p,
  width = 6,
  height = 5,
  units = "in",
  dpi = 600,
  compression = "lzw"
)
```





Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
